# Construction of data set "berlin.rda"

## download and unzip the data

```{r}
u <- "https://www.vbb.de/media/download/2029"
utils::download.file(u, "./data-raw/vbb.zip")
unzip ("vbb.zip", exdir = "./data-raw/")
```

## Reduce down to U and S Bahn only and recreate vbb.zip

First remove GTFS files that are not directly needed for routing
```{r}
rms <- c ("./data-raw/shapes.txt", "./data-raw/agency.txt",
          "./data-raw/calendar_dates.txt", "./data-raw/calendar.txt")
invisible (file.remove (rms))
```

Then cut down the data only to S and U bahn routes and stops, with `stop_times`
reduced to the single hour between 11:00-12:00. Non-ASCII characters also have
to be converted to pass checks (so all umlauts and eszetts are removed).
```{r}
library (stringi)
routes <- data.table::fread ("./data-raw/routes.txt", integer64 = "character")
index_s <- grep ("^S", routes [, route_short_name])
index_u <- grep ("^U", routes [, route_short_name])
routes <- routes [c (index_s, index_u), ]
data.table::fwrite (routes, "./data-raw/routes.txt", quote = TRUE)

trips <- data.table::fread ("./data-raw/trips.txt", integer64 = "character")
trips <- trips [which (trips [, route_id] %in% routes [, route_id]), ]
trips <- trips [, trip_headsign := stri_trans_general (trip_headsign, "latin-ascii")]
data.table::fwrite (trips, "./data-raw/trips.txt", quote = TRUE)

stop_times <- data.table::fread ("./data-raw/stop_times.txt",
                                 integer64 = "character", showProgress = FALSE)
stop_times <- stop_times [which (stop_times [, trip_id] %in% trips [, trip_id]), ]
stop_times <- stop_times [, stop_headsign :=
                          stri_trans_general (stop_headsign, "latin-ascii")]
arrival_time <- gtfsrouter:::rcpp_time_to_seconds (stop_times [, arrival_time])
departure_time <- gtfsrouter:::rcpp_time_to_seconds (stop_times [, departure_time])
index <- which (departure_time > 12 * 3600 & arrival_time < 13 * 3600)
data.table::fwrite (stop_times [index, ], "./data-raw/stop_times.txt", quote = TRUE)

stops <- data.table::fread ("./data-raw/stops.txt",
                            integer64 = "character", showProgress = FALSE)
stops <- stops [which (stops [, stop_id] %in% stop_times [, stop_id]), ]
stops <- stops [, stop_name := stri_trans_general (stop_name, "latin-ascii")]
data.table::fwrite (stops, "./data-raw/stops.txt", quote = TRUE)

transfers <- data.table::fread ("./data-raw/transfers.txt",
                                integer64 = "character", showProgress = FALSE)
index <- which ((transfers [, from_stop_id] %in% stops [, stop_id]) &
                (transfers [, to_stop_id] %in% stops [, stop_id]))
transfers <- transfers [index, ]
data.table::fwrite (transfers, "./data-raw/transfers.txt", quote = TRUE)
```


Re-zip
```{r}
flist <- c ("./data-raw/routes.txt", "./data-raw/trips.txt", "./data-raw/stop_times.txt",
            "./data-raw/stops.txt", "./data-raw/transfers.txt")
zip ("./data-raw/vbb.zip", files = flist)
invisible (file.remove (flist))
```

## Then extract contents for test data

```{r}
unzip ("./data-raw/vbb.zip", exdir = ".")
berlin_gtfs <- list (routes = data.table::fread ("./data-raw/routes.txt",
                                                 integer64 = "character"),
              trips = data.table::fread ("./data-raw/trips.txt",
                                         integer64 = "character"),
              stop_times = data.table::fread ("./data-raw/stop_times.txt",
                                              integer64 = "character"),
              stops = data.table::fread ("./data-raw/stops.txt",
                                         integer64 = "character"),
              transfers = data.table::fread ("./data-raw/transfers.txt",
                                             integer64 = "character"))
usethis::use_data (berlin_gtfs, overwrite = TRUE, compress = "xz")
flist <- c ("./data-raw/routes.txt", "./data-raw/trips.txt", "./data-raw/stop_times.txt",
            "./data-raw/stops.txt", "./data-raw/transfers.txt")
invisible (file.remove (flist))
invisible (file.remove ("./data-raw/vbb.zip"))
```

