---
title: "gtfsrouter"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: flatly
header-includes: 
    - \usepackage{tikz}
    - \usetikzlibrary{arrows}
vignette: >
  %\VignetteIndexEntry{gtfsrouter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
options (warn = 1)
```


```{r pkg-load, echo = FALSE, message = FALSE}
library (gtfsrouter)
#devtools::load_all (".", export_all = FALSE)
```

# 1 Background: GTFS and other R packages

GTFS stands for General Transit Feed Specification. It began life in 2005 as the
"Google Transit Feed Specification," which was renamed to "General" in 2009.  A
GTFS data set consists of several comma-delimited (`.csv`) files detailing
routes, stops, trips, transfers, and other aspects, all bundled in a single
`.zip`-compressed archive file. For full details, see the relevant [google
developer site](https://developers.google.com/transit/gtfs/).

There are currently two other **R** packages which handle GTFS data:

1. [`gtfsr`](https://github.com/ropensci/gtfsr), hosted by
   [rOpenSci](https://ropensci.org), developed by [Danton
   Noriega](https://github.com/dantonnoriega), but no longer under active
   development.
2. [`tidytransit`](https://github.com/r-transit/tidytransit), which began as a
   fork of [`gtfsr`](https://github.com/ropensci/gtfsr), and currently
   represents its successor. This package can be used to, "map transit stops and
   routes, calculate transit frequencies, and validate transit feeds [as well as
   to read] the General Transit Feed Specification into tidyverse and simple
   features dataframes."

The one thing neither of these packages enable is the use of GTFS data for
transit routing, which is where the `gtfsrouter` package comes in. Beyond
enabling A-to-B routing, the package also provides various functions for the
statistical analysis of GTFS data sets. It does this through converting the full
timetable contained in a GTFS data set into a so-called "median timetable"
representing the median service frequencies at every stop, and the median travel
times between each pair of stops. This vignette is accordingly divided into two
major sections: routing and median timetables, respectively describing functions
for A-to-B routing, and larger-scale statistical analyses.

Functionality in both cases is demonstrated through the sample data set included
with the package, provided by the "Verkehrsverbund Berlin-Brandenburg" (VBB; or
Transport Network Berlin-Brandenburg). The `berlin_gtfs` data represents a
reduced version of the full GTFS data, containing only six tables, and a
timetable reduced to the single hour between 12:00-13:00. Like all GTFS
software including [`tidytransit`](https://github.com/r-transit/tidytransit),
this package is designed to work directly with GTFS data in `.zip`-archived
format, and so includes a helper function, `berlin_gtfs_to_zip()`, which exports
the internal data set to a locally-stored `.zip` archive in the `tempdir()` of
the current **R** session. These data can be exported and re-imported with:
```{r berlin_gtfs}
berlin_gtfs_to_zip()
f <- file.path (tempdir (), "vbb.zip")
file.exists (f)
gtfs <- extract_gtfs (f)
```
That simply re-creates the original package data, `berlin_gtfs` (although the
extracted data differ through having a couple of additional attributes defining
it as a `gtfs` object).
     



# 2. Routing

The primary routing function is `gtfs_route()`, the example of which uses the
`gtfs` data for the VBB created as described above. In simplest form, routing
requires a start and end point, defaulting to the current time as desired start
time, and routing for the current day of the week.
```{r route1-fakey, eval = FALSE}
from <- "Innsbrucker Platz"
to <- "Alexanderplatz"
gtfs_route (gtfs, from = from, to = to)
```
```{r route1, eval = TRUE, echo = FALSE}
from <- "Innsbrucker Platz"
to <- "Alexanderplatz"
knitr::kable (gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00"))
```

Both the start time and day of the week can be explicitly specified:
```{r route2, eval = TRUE}
route <- gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday")
```

## 2.1 GTFS Timetables

The `gtfsrouter` package uses the [Connection Scan
Algorithm](https://arxiv.org/abs/1703.05997), which requires converting the
"stop_times" table to a column-wise timetable. The "stop_times" table has
row-wise entries for each distinct "trip_id", with consecutive rows for a given
value of "trip_id" holding sequential values for stops and associated times (and
potentially additional variables). In contrast, the timetables processed by this
package have separate columns for departure and arrival stations and times. All
routing queries pre-process the original GTFS data with the `gtfs_timetable()`
function, which appends this timetable data, along with two single-column tables
of stop and trip ID values. (The timetable itself contains strictly integer
values for stops and trips, which are indices into these latter tables.)

The only important point of that from a user's perspective is that routing
queries will be faster if this pre-processing step is explicitly implemented
with `gtfs_timetable()` prior to calling `gtfs_route()`. This is easy to
demonstrate using the sample data:
```{r timetable}
gtfs <- extract_gtfs (f)
from <- "Innsbrucker Platz"
to <- "Alexanderplatz"
system.time (gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday"))
names (gtfs)
gtfs <- gtfs_timetable (gtfs, day = "Sunday") # explicit pre-processing to extract timetable for Sunday
names (gtfs)
system.time (gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00"))
```
Note that the `day` parameter is used to extract the timetable, after which it
is no longer required in the actual call to `gtfs_route()`.


## 2.1. Routing by mode of transport

It is also possible to filter by desired mode of transport. This is done by
matching the pattern to those given in the `route_short_name` column of the
`gtfs$route` table:
```{r gtfs_route_table-fakey, eval = FALSE}
head (gtfs$route)
```
```{r gtfs_route_table, echo = FALSE}
knitr::kable (head (gtfs$route))
```

These short names will differ for each GTFS, with the two primary train systems
in Berlin being the underground trains denoted "U" (although not always
travelling underground), and street-level trains denoted "S". The default route
from `r from` to `r to` above was via two "U" services.  We can also specify
that we'd prefer to travel by "S" services, noting that the `route_pattern =
"S"` specifies a `route_short_name` that *starts with* (`"^"`) "S":
```{r route3-fakey, eval = FALSE}
gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday", route_pattern = "^S")
```
```{r route3, echo = FALSE, eval = TRUE}
knitr::kable (gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday", route_pattern = "^S"))
```

## 2.2. Routing for earliest arrivals or earliest departures

The above route with the "S" services leaves one minute later, and arrives two
minutes later. Importantly, `gtfs_route()` searches by default for the service
which arrives at the nominated destination station at the earliest time. This
may not always be the first available service departing from the nominated start
station. Routing with the earliest *departing* service, instead of the earliest
*arriving* service, can be specified with the binary `earliest_arrival`
parameter:
```{r route4a-fakey, eval = FALSE}
from <- "Alexanderplatz"
to <- "Pankow"
gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday", earliest_arrival = FALSE)
```
```{r route4a, eval = TRUE, echo = FALSE}
from <- "Alexanderplatz"
to <- "Pankow"
r1 <- gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday")
r2 <- gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday", earliest_arrival = FALSE)
knitr::kable (r2)
```

And the earliest-departing route arrives at `r to` at 
`r r2$arrival_time [nrow (r2)]`, departing `r from` at 
`r r2$departure_time [1]`. In contrast, the earliest-arriving service is:
```{r route4b-fakey, eval = FALSE}
gtfs_route (gtfs, from = from, to = to, start_time = "12:00:00", day = "Sunday", earliest_arrival = TRUE)
```
```{r route4b, eval = TRUE, echo = FALSE}
knitr::kable (r1)
diff_arrival <- gtfsrouter:::convert_time (r2$arrival_time [nrow (r2)]) -
    gtfsrouter:::convert_time (r1$arrival_time [nrow (r1)])
diff_depart <- gtfsrouter:::convert_time (r1$departure_time [1]) -
    gtfsrouter:::convert_time (r2$departure_time [1])
mm <- floor (diff_depart / 60)
ss <- diff_depart - mm * 60
diff_depart <- paste0 (mm, "min, ", ss, "s")

diff_total <- (gtfsrouter:::convert_time (r2$arrival_time [nrow (r2)]) -
    gtfsrouter:::convert_time (r2$departure_time [1])) -
    (gtfsrouter:::convert_time (r1$arrival_time [nrow (r1)]) -
     gtfsrouter:::convert_time (r1$departure_time [1]))
mm <- floor (diff_total / 60)
ss <- diff_total - mm * 60
diff_total <- paste0 (mm, "min, ", ss, "s")
```

This service departs `r diff_depart` later at `r r1$departure_time [1]`, and
arrives `r diff_arrival` seconds earlier at `r r1$arrival_time [nrow (r1)]`. The
earliest-arriving service thus entails `r diff_total` less travel time than the
earliest departing service.  It is nevertheless important to note that queries
for earliest-arriving services require two full routing runs, whereas
earliest-departing services can be executed in a single run. This, bulk queries
for analytic purposes will generally be up to twice as first with
`earliest_arrival = FALSE`.


# 3. Median Timetables

```{r}
options (warn = 0)
```
